AGB Stars in the IR: Physical Characteristics, Galactic Spatial Distribution, and Chemical Impact of the Milky Way AGB Population

### 2015-11-13 paper1/paper1_data.tex
Everything to the end of section 3

Our principal data source is the merged WISE-2MASS catalog. In order to derive
WISE-2MASS color-based selection criteria, we use a homogenous sample of AGB stars from the OGLE-III Catalog of Variable Stars supplying objects in the Magellanic Clouds. This sample serves to calibrate color-color and color-absolute magnitude
relations that underlie our distance estimates. In order to assess what infrared populations
could contribute to contamination of selected \agb\ candidates, we utilize
extragalactic sources from SDSS data release 7 pulled from the NYU Value-added Galaxy Catalog,
the SDSS unofficial Luminous Red Galaxy sample,
and young stellar objects identified in \wise\ data, also described below. 
We conclude this section by describing how these auxiliary catalogs were positionally merged
with the WISE-2MASS catalog, summarize the object counts before and after quality control (Table~\ref{tab:reductions}), and summarize the effects of WISE-2MASS color-color cuts (Table~\ref{tab:criteria_completeness_contamination}) on the AGB and contaminant samples.

@@

\subsection{AGB stars from the OGLE-III Catalog of Variable Stars}\label{sec:ogle}
The OGLE-III Catalog of Variable Stars (CVS) \citep{2008AcA....58...69U,2009AcA....59..239S,2011AcA....61..217S} is a subset of the overall OGLE-III dataset, containing roughly 10 years of observations in the $V$- and $I$-bands of over 120,000 variable stars, with typically 700-900 data points per star in the $I$-band and about 50 data points per star in $V$-band. With high-precision photometry \citep[$\sim0.01$ mag,][]{2007AcA....57..201S}, these observations saturate at $I = 12.5$ mag, and are limited at the faint end at $I = 20.5$ mag \citep{2001AcA....51..303Z}. The variability of these stars covers a wide range of periodicity ($4 < P < 2000$ days) and amplitudes ($0.005 < A_I < 5.7$ mag).

The OGLE-III CVS is accessible through an online database\footnote{\url{http://ogledb.astrouw.edu.pl/~ogle/CVS/}}, with long-period variables (LPVs) classified by object type, evolutionary status, and spectral type. Object type is based chiefly on variability and luminosity. OGLE Small Amplitude Red Giants (hereon OSARGs) are weakly-variable ($0.005 < A_I < 0.13$ mag), with relatively short periods \citep[$10 < P < 100$ days;][]{2004AcA....54..129S}. Additionally, because these objects are multiperiodic, OSARGs are selected using their period ratios and their distances from established period-luminosity (log$P-L$) sequences from \cite{1999IAUS..191..151W}. For a more detailed description of their selection, see \cite{2007AcA....57..201S}. Mira variables and Semi-Regular Variables (hereon SRVs) are identified using $I$-band amplitudes and $P-W_I$, where $W_I$ is the reddening-free Wesenheit index \citep{2005AcA....55..331S}:

@@

As red giant branch stars (RGBs) can contaminate the same log($P)-L$ space as AGB stars, evolutionary type is distinguished in two ways. Stars above the tip of the RGB ($K_s = 12.05$ mag in LMC; \citealt{2004AcA....54..129S, 2007AcA....57..201S}) are classified as AGBs. Below that threshold, a narrow sequence of stars appears to share the same slope and intercept with AGB stars above the TRGB. Additionally they share similar primary-secondary period ratios as known AGB stars \citep{2005AcA....55..331S}. However, as there is still some contamination by RGBs using period ratios, we reject all OGLE objects below the TRGB.

@@

From the initial OGLE-III sample of 52,976 LPVs, we retain 43,262 after matching to ALLWISE within 3", and ensuring only 1 match to 2MASS as well as all objects needing to be in the LMC. The $I$-band median of OGLE-III LPVs is $\approx14.68$ mag with a standard deviation of 0.72 mag. Out of the entire sample only 13 objects (Miras) fall beyond the 20.5 mag 5$\sigma$ faint limit, and 41 beyond the saturation limit. As such, we expect this to effectively represent the complete sample of oscillating AGB stars in the LMC.

Out of the 43,262 objects matched to ALLWISE within 3", 1,633 are Mira variables, 30,768 are OSARGs, and 10,861 are SRVs. The right side of Figure~\ref{fig:oglepmag} shows color-color plots for OGLE-2MASS colors, with a line at $(J-K_s) = 1.4$ drawn to separate O-rich and C-rich Miras. Out of the total of 457 O-rich Miras, 36 (7.9\%) have $(J-K_s) > 1.4$. Conversely, out of the 1,176 C-rich total Miras, 1,138 (96.8\%) have $(J-K_s) > 1.4$. The distribution of these objects on the sky is summarized in Figure~\ref{fig:oglegalplot}.

@@

\subsection{WISE+2MASS Stellar Locus}
The color-color space occupied by AGB stars showing significant photospheric emission is heavily populated by naked stars from the main sequence. By sheer number, these stars would drown out sources that could be AGB stars with thin circumstellar envelopes. To address this, we extract the stellar locus from ALLWISE in the direction of the LMC ($276.5^\circ < l < 284^\circ$, $-38.2^\circ < b < -28^\circ$). We require that every object have 1 2MASS association, [$W1/W2/W3$] signal-to-noise $>$ 1, detections in 2MASS J, H, and K,  and confusion \& contamination flags set to 0 for [$W1/W2/W3$]. We then adapt the color-color criteria from \cite{2014MNRAS.440.3430D}. Their locus focuses on stars with effective temperatures $3540 < T_\text{eff} < 7200 K$. We fit their criteria for $J-K_s$ vs $W1 - W2$ within 3$\sigma$ of the locus with degree-3 polynomials. The resulting fit for the 3$\sigma$ stellar locus follows:

@@

\begin{figure}[h]
\includegraphics[width=3in]{figs/lmc_locus_3sig.png}
\caption{The WISE-2MASS stellar locus. (\emph{black dots}) objects from the stellar locus in the direction of the LMC. (\emph{blue line}) WISE-2MASS color-color stellar locus derived from \cite{2014MNRAS.440.3430D}. (\emph{red line}) 3$\sigma$ boundaries for WISE-2MASS color-color locus. \label{fig:locus}}
\end{figure}

@@

The final OGLE-2MASS-WISE AGB sample is produced from the obejct sample in section~\ref{sec:ogle}. We apply the point source saturation limits of [$W1/W2/W3/K_s$] $<$ [$2.0/1.5/-3.0/8.5$]. We also enforce, in order, the 5$\sigma$ faint limits for [$W1/W2/W3/K_s$]  $<$ [$16.83/15.6/11.32/15.5$]. The WISE point source saturation and 5$\sigma$ faint limits can be found in \cite{2012wise.rept....1C}. The 2MASS faint and saturation limits are found in \cite{2006AJ....131.1163S}. To ensure quality measurements in the desired bands, we also enforce [$W1/W2/W3$] SNR $> 3$, and contamination \& confusion flags ({\tt cc\_flag}) set to 0 in each of those bands. The VAGC, LRG, and YSO samples are filtered through the same criteria. Note that we deliberately neglect $W4$ because, after applying the 5$\sigma$ faint limits as well as {\tt cc\_flag}=0, our sample is too heavily reduced. For similar reasons, we only apply the {\tt cc\_flag} criteria to [$W1/W2$], noting that clean flags in these two bands should be sufficient for source detection considering the extended NEOWISE mission contains extra observations for these wavelengh regimes. The resulting populations from each reduction step can be found in Table~\ref{tab:reductions}.

@@

	\scalebox{0.85}{
		\begin{tabular}{l r r r r r r r r r r}
		\hline\hline
		Population & OGLE & QSO & AGN & SF & SB & LRGs & YSOs & Locus & \textbf{Total} \\ 
		\hline
		Original & 46,262 & 122,550 & 19,184 & 232,845 & 67,128 & 105,631 & 290 & 25,254 & $\mathbf{619,267}$ \\
		WISE match & 43,209 & 6,902 & 3,098 & 36,539 & 10,359 & 75,543 & 274 & 25,254 & $\mathbf{201,178}$ \\
		$W1$-3, $K_s$ saturation limit& 43,201 & 6,900 & 3,098 & 36,538 & 10,358 & 75,527 & 214 & 25,156 & $\mathbf{200,992}$ \\
		 $W1$-2 faint limit & 43,200 & 6,891 & 3,098 & 36,537 & 10,358 & 75,527 & 214 & 24,382 & $\mathbf{200,207}$  \\
		 $W1$-3 faint limit & 19,358 & 6,728 & 3,005 & 35,678 & 10,105 & 1,184 & 214 & 3,154 & $\mathbf{79,426}$  \\
		 $W1$-3, $K_s$ faint limit & 19,358 & 5,283 & 2,878 & 34,299 & 9,676 & 1,167 & 213 & 3,070 & $\mathbf{75,944}$ \\
		 $W1$-3 SNR $> 3$ & 18,995 & 5,279 & 2,876 & 34,285 & 9,673 & 1,079 & 213 & 3,056 & $\mathbf{75,456}$\\
		 cc\_flag = 0 & 9,239 & 5,029 &  2,741 &  32,805 &  9,219 &  1,038 & 184 &  3,056 & $\mathbf{63,308}$\\
		\hline\hline
		\end{tabular}}
	\caption{Sample populations with respect to each reduction step. Please see section~\ref{sec:merged} for the appropriate saturation limits, faint limits, and WISE matching radii.\label{tab:reductions}}
	\end{center}

>>>>

Our principal data source is the merged WISE-2MASS catalog. In order to derive
WISE-2MASS color-based selection criteria, we use a homogenous sample of AGB stars from the OGLE-III Catalog of Variable Stars supplying objects in the Magellanic Clouds. This sample serves to calibrate color-color and color-absolute magnitude
relations that underlie our distance estimates. In order to assess what infrared populations
could contribute to contamination of selected \agb\ candidates, we utilize stellar and 
extragalactic sources from SDSS data release 7 pulled from the NYU Value-added Galaxy Catalog,
the SDSS unofficial Luminous Red Galaxy sample, young stellar objects identified in \wise\ data, and red giant branch stars from the OGLE-III Catalog of Variable Stars, also described below. 
We conclude this section by describing how these auxiliary catalogs were positionally merged
with the WISE-2MASS catalog, summarize the object counts before and after quality control (Table~\ref{tab:reductions}), and summarize the effects of WISE-2MASS color-color cuts (Table~\ref{tab:criteria_completeness_contamination}) on the AGB and contaminant samples.

@@

\subsection{Variable Giants from the OGLE-III Catalog of Variable Stars}\label{sec:ogle}
The OGLE-III Catalog of Variable Stars (CVS) \citep{2008AcA....58...69U,2009AcA....59..239S,2011AcA....61..217S} is a subset of the overall OGLE-III dataset, containing roughly 10 years of observations in the $V$- and $I$-bands of over 120,000 variable stars, with typically 700-900 data points per star in the $I$-band and about 50 data points per star in $V$-band. With high-precision photometry \citep[$\sim0.01$ mag,][]{2007AcA....57..201S}, these observations saturate at $I = 12.5$ mag, and are limited at the faint end at $I = 20.5$ mag \citep{2001AcA....51..303Z}. The variability of these stars covers a wide range of periodicity ($4 < P < 2000$ days) and amplitudes ($0.005 < A_I < 5.7$ mag).

The OGLE-III CVS is accessible through an online database\footnote{\url{http://ogledb.astrouw.edu.pl/~ogle/CVS/}}, with long-period variables (LPVs) classified by object type, evolutionary status , and spectral type. 
\subsubsection{AGB Stars}
Stars brighter than the tip of the red giant branch (RGB) ($K_s < 12.05$ mag in LMC; \citealt{2004AcA....54..129S, 2007AcA....57..201S}) are classified as AGBs. Object types (OSARG, Mira, SRV) are based chiefly on variability and luminosity. OGLE Small Amplitude Red Giants (hereon OSARGs) are weakly-variable ($0.005 < A_I < 0.13$ mag), with relatively short periods \citep[$10 < P < 100$ days;][]{2004AcA....54..129S}. Additionally, because these objects are multiperiodic, OSARGs are selected using their period ratios and their distances from established period-luminosity (log$P-L$) sequences from \cite{1999IAUS..191..151W}. For a more detailed description of their selection, see \cite{2007AcA....57..201S}. Mira variables and Semi-Regular Variables (hereon SRVs) are identified using $I$-band amplitudes and $P-W_I$, where $W_I$ is the reddening-free Wesenheit index \citep{2005AcA....55..331S}:

@@

@@

From the initial OGLE-III LMC sample of 46,467 LPVs we retain 44,013 after matching to ALLWISE within 3", and 43,247 after ensuring only 1 match to 2MASS. The $I$-band median of OGLE-III LPVs is $\approx14.68$ mag with a standard deviation of 0.72 mag, appearing well above the OGLE $I$-band 5$\sigma$ faint limit. Out of the entire sample only 13 objects (Miras) fall below the 5$\sigma$ faint limit, and 41 above the saturation limit. Because so few stars even approach the faint limit of OGLE, we expect the OGLE-2MASS catalog of AGBs to effectively represent the complete sample of AGB stars in the LMC down to $I < 20.5$, oscillating with amplitudes $> 0.005$ mag.

Out of the 43,247 objects matched to ALLWISE and 2MASS within 3", 1,633 are Mira variables, 30,756 are OSARGs, and 10,858 are SRVs. %The right side of Figure~\ref{fig:oglepmag} shows color-color plots for OGLE-2MASS colors, with a line at $(J-K_s) = 1.4$ drawn to separate O-rich and C-rich Miras. Out of the total of 457 O-rich Miras, 36 (7.9\%) have $(J-K_s) > 1.4$. Conversely, out of the 1,176 C-rich total Miras, 1,138 (96.8\%) have $(J-K_s) > 1.4$. 
The distributions of these objects in log$(P)-L$ and visible-NIR color-color space are summarized in Figure~\ref{fig:oglegalplot}.

\subsubsection{Red Giant Branch Stars}
As RGBs can contaminate the same log($P)-L$ space and color-color space as AGB stars, we seek to use RGBs identified by the OGLE-III CVS to quantify their contamination of our AGB sample. Stars brighter than the tip of the RGB ($K_s < 12.05$ mag in LMC; \citealt{2004AcA....54..129S, 2007AcA....57..201S}) are classified as AGBs. Below that threshold, narrow sequences of variable stars appear to share similar slopes and intercept with AGB stars above the TRGB. \cite{2004AcA....54..129S} classifies these objects as RGBs. The OGLE-III CVS LMC sample initially contains 45,528 RGB stars. When matched to ALLWISE-2MASS catalog within 3" we retain 40,764 RGBs.

@@

\subsection{WISE+2MASS Stellar Locus}
The color-color space occupied by AGB stars showing significant photospheric emission in WISE filters is heavily populated by naked stars from the main sequence \citep{2014MNRAS.442.3361N}. By sheer number, these stars would drown out sources that could be AGB stars with thin circumstellar envelopes. To address this, we extract the stellar locus from ALLWISE in the direction of the LMC ($276.5^\circ < l < 284^\circ$, $-38.2^\circ < b < -28^\circ$). We require that every object have one 2MASS association, [$W1/W2/W3$] signal-to-noise $>$ 1, detections in 2MASS $J$, $H$, and $K_s$,  and confusion \& contamination flags set to 0 for [$W1/W2/W3$]. We then adapt the color-color criteria from \cite{2014MNRAS.440.3430D}. Their locus focuses on stars with effective temperatures $3540 < T_\text{eff} < 7200 K$. We fit their criteria for $J-K_s$ vs $W1 - W2$ within 3$\sigma$ of the locus with degree-3 polynomials. The resulting fit for the 3$\sigma$ stellar locus follows:

@@

\begin{figure}[h]
\includegraphics[width=3in]{figs/lmc_locus_3sig.png}
\caption{The WISE-2MASS stellar locus. (\emph{black dots}) objects from the stellar locus in the direction of the LMC. (\emph{green line}) WISE-2MASS color-color stellar locus derived from \cite{2014MNRAS.440.3430D}. (\emph{red line}) 3$\sigma$ boundaries for WISE-2MASS color-color locus. \label{fig:locus}}
\end{figure}

@@

The final OGLE-2MASS-WISE AGB sample is produced from the object sample in section~\ref{sec:ogle}. We apply the point source saturation limits of [$W1/W2/W3/K_s$] $<$ [$2.0/1.5/-3.0/8.5$]. We also enforce, in order, the 5$\sigma$ faint limits for [$W1/W2/W3/K_s$]  $<$ [$16.83/15.6/11.32/15.5$]. The WISE point source saturation and 5$\sigma$ faint limits can be found in \cite{2012wise.rept....1C}. The 2MASS faint and saturation limits are found in \cite{2006AJ....131.1163S}. To ensure quality measurements in the desired bands, we also enforce [$W1/W2/W3$] SNR $> 3$, and contamination \& confusion flags ({\tt cc\_flag}) set to 0 in each of those bands. The VAGC, LRG, and YSO samples are filtered through the same criteria. Note that we deliberately neglect $W4$ because, after applying the 5$\sigma$ faint limits as well as {\tt cc\_flag}=0, our sample is too heavily reduced. %For similar reasons, we only apply the {\tt cc\_flag} criteria to [$W1/W2$], noting that clean flags in these two bands should be sufficient for source detection considering the extended NEOWISE mission contains extra observations for these wavelengh regimes. 
The resulting populations from each reduction step can be found in Table~\ref{tab:reductions}.

@@

	\scalebox{0.85}{
		\begin{tabular}{l r r r r r r r r r r}
		\hline\hline
		Population & OGLE & QSO & AGN & SF & SB & LRGs & YSOs & Locus & RGBs & \textbf{Total} \\ 
		\hline
		Original & 46,467 & 122,550 & 19,184 & 232,845 & 67,128 & 105,631 & 290 & 25,254 & 45,528 & $\mathbf{664,877}$ \\
		WISE match & 43,247 & 6,902 & 3,098 & 36,539 & 10,359 & 75,543 & 274 & 25,254 & 40,764 & $\mathbf{241,980}$ \\
		$W1$-3, $K_s$ sat. limit& 43,201 & 6,900 & 3,098 & 36,538 & 10,358 & 75,527 & 214 & 25,156 & 40,763 & $\mathbf{241,755}$ \\
		 $W1$-2 faint limit & 43,200 & 6,891 & 3,098 & 36,537 & 10,358 & 75,527 & 214 & 24,382 & 40,760 & $\mathbf{240,967}$  \\
		 $W1$-3 faint limit & 19,358 & 6,728 & 3,005 & 35,678 & 10,105 & 1,184 & 214 & 3,154 & 5,170 & $\mathbf{84,596}$  \\
		 $W1$-3, $K_s$ faint limit & 19,358 & 5,283 & 2,878 & 34,299 & 9,676 & 1,167 & 213 & 3,070 & 5,170 & $\mathbf{81,114}$ \\
		 $W1$-3 SNR $> 3$ & 18,995 & 5,279 & 2,876 & 34,285 & 9,673 & 1,079 & 213 & 3,056 & 4,727 & $\mathbf{80,183}$\\
		 cc\_flag = 0 & 8,454 & 5,029 &  2,741 &  32,805 &  9,219 &  1,038 & 174 &  3,056 & 1,599 & $\mathbf{64,125}$\\
		\hline\hline
		\end{tabular}}
	\caption{Sample populations with respect to each reduction step. Please see section~\ref{sec:merged} for the appropriate saturation limits, faint limits, and WISE matching radii.\label{tab:reductions}}
	\end{center}
